Parallel Hierarchical Clustering using \\ Rank-Two Nonnegative Matrix Factorization \thanks{Identify applicable funding agency here. If none, delete this.}

### 2020-07-03 latex/fig/tree.tex

\centering
\begin{tikzpicture}[scale=.9,level/.style={sibling distance=50mm/#1}]
\node [circle,draw] (z) {}
  child {node [circle,draw,fill=gray] (a) {}
    child {node [circle,draw,dashed] (b) {}}
    child {node [circle,draw,dashed] (g) {}}
  }
  child {node [circle,draw] (j) {}
    child {node [circle,draw] (k) {}
      child {node [circle,draw,fill=gray] (o) {}
        child {node [circle,draw,dashed] (x) {}}
        child {node [circle,draw,dashed] (y) {}}
      }
      child {node [circle,draw,fill=gray] (p) {}
        child {node [circle,draw,dashed] (x) {}}
        child {node [circle,draw,dashed] (y) {}}
      }
    }
  child {node [circle,draw,fill=gray] (l) {}
    child {node [circle,draw,dashed] (c) {}}
    child {node [circle,draw,dashed] (d) {}}
  }
};

>>>>

\centering
\begin{tikzpicture}[scale=.9,level/.style={sibling distance=50mm/#1}]
\node [circle,draw] (z) {}
  child {node [circle,draw,fill=gray] (a) {}
    child[dashed] {node [circle,draw,dashed] (b) {}}
    child[dashed] {node [circle,draw,dashed] (g) {}}
  }
  child {node [circle,draw] (j) {}
    child {node [circle,draw] (k) {}
      child {node [circle,draw,fill=gray] (o) {}
        child[dashed] {node [circle,draw,dashed] (x) {}}
        child[dashed] {node [circle,draw,dashed] (y) {}}
      }
      child {node [circle,draw,fill=gray] (p) {}
        child[dashed] {node [circle,draw,dashed] (x) {}}
        child[dashed] {node [circle,draw,dashed] (y) {}}
      }
    }
  child {node [circle,draw,fill=gray] (l) {}
    child[dashed] {node [circle,draw,dashed] (c) {}}
    child[dashed] {node [circle,draw,dashed] (d) {}}
  }
};

### 2020-07-03 latex/paper.tex
added a few notes

@@

\begin{itemize}
	\item NMF, BCD, rank-2 NMF
\end{itemize}

@@

\begin{algorithm}
\caption{Rank-2 Nonnegative Least Squares Solve \cite{KP13}}
\label{alg:r2nmf}
\begin{algorithmic}[1]
	\Require{$\M{C}$ is $n\times 2$ and $\M{G}$ is $2\times 2$ and s.p.d.}
	\Function{$\M{H} =$ Rank2-NLS-Solve}{$\M{C},\M{G}$}
		\State $\M{H} = \M{C}\M{G}^{-1}$
		\For{$i=1$ to $n$}
			\If{$\ME{H}{i1} < 0$ or $\ME{H}{i2} < 0$}
				\If{$\ME{H}{i1} \sqrt{\ME{G}{11}}  < \ME{H}{i2} \sqrt{\ME{G}{22}} $}
					\State $\ME{H}{i1} = 0$
					\State $\ME{H}{i2} = \ME{C}{i2} / \ME{G}{22}$
				\Else
					\State $\ME{H}{i1} = \ME{C}{i1} / \ME{G}{11}$
					\State $\ME{H}{i2} = 0$
				\EndIf
			\EndIf
		\EndFor
	\EndFunction
	\Ensure{$\displaystyle \M{H} = \argmin_{\M[\bar]{H} \geq \M{0}} \|\M{A}-\M{W}\M[\bar]{H}^\Tra\|$ is $n\times 2$ with $\M{C}=\M{A}^\Tra \M{W}$ and $\M{G}=\M{W}^\Tra\M{W}$}
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
\caption{Hierarchical Clustering \cite{KP13}}
\label{alg:hiernmf2}
\begin{algorithmic}[1]
	\Require{$\M{A}$ is $m\times n$, $k$ is target number of leaves}
	\Function{${\cal T} =$ Hier-R2-NMF}{$\M{A}$}
		\State ${\cal R} = \text{node}(\M{A})$ \hfill \Comment{create root node}
		\State \textsc{Split}$({\cal R})$
		\State inject$({\cal Q},{\cal R}.\text{left})$ \hfill \Comment{create priority queue}
		\State inject$({\cal Q},{\cal R}.\text{right})$ \hfill \Comment{of frontier nodes}
		\While{size$({\cal Q}) < k$}
			\State ${\cal N} = \text{eject}({\cal Q})$ \hfill \Comment{get leaf with max score}
			\State \textsc{Split}$({\cal R}.\text{left})$ \hfill \Comment{split left child}
			\State inject$({\cal Q},{\cal R}.\text{left})$ \hfill \Comment{and add to $\cal Q$}
			\State \textsc{Split}$({\cal R}.\text{right})$ \hfill \Comment{split right child}
			\State inject$({\cal Q},{\cal R}.\text{right})$ \hfill \Comment{and add to $\cal Q$}
		\EndWhile
	\EndFunction
	\Ensure{$\cal{T}$ is binary tree rooted at $\cal R$ with $k$ frontier nodes, each node has subset of cols of $\M{A}$ and feature vector $\V{w}$}
\end{algorithmic}
\end{algorithm}

>>>>

\GB{ask Haesun to write intro paragraph motivating hierarchical clustering but noting the limitations of time and data size}

@@

\begin{itemize}
	\item NMF, BCD, rank-2 NMF
	\item PLANC
\end{itemize}

@@

\begin{itemize}
	\item subroutine for splitting node is rank-2 NMF
	\item using BCD, subproblem is set of independent NNLS problems with variable vector of length 2
	\item 4 possibilities
	\begin{itemize}
		\item all-zero option never optimum if input is nonnegative
		\item if all-pos option is optimum, it is solution to unconstrained problem
		\item can choose between one-zero options using line 6 of algorithm
		\item describe \cref{alg:r2nmf}
	\end{itemize}
\end{itemize}

\begin{algorithm}
\caption{Rank-2 Nonnegative Least Squares Solve \cite{KP13}}
\label{alg:r2nmf}
\begin{algorithmic}[1]
	\Require{$\M{C}$ is $n\times 2$ and $\M{G}$ is $2\times 2$ and s.p.d.}
	\Function{$\M{H} =$ Rank2-NLS-Solve}{$\M{C},\M{G}$}
		\State $\M{H} = \M{C}\M{G}^{-1}$ \hfill \Comment{Solve unconstrained system}
		\For{$i=1$ to $n$}
			\If{$\ME{H}{i1} < 0$ or $\ME{H}{i2} < 0$}
				\State \Comment{Choose between single-variable solutions}
				\If{$\ME{H}{i1} \sqrt{\ME{G}{11}}  < \ME{H}{i2} \sqrt{\ME{G}{22}} $}
					\State $\ME{H}{i1} = 0$
					\State $\ME{H}{i2} = \ME{C}{i2} / \ME{G}{22}$
				\Else
					\State $\ME{H}{i1} = \ME{C}{i1} / \ME{G}{11}$
					\State $\ME{H}{i2} = 0$
				\EndIf
			\EndIf
		\EndFor
	\EndFunction
	\Ensure{$\displaystyle \M{H} = \argmin_{\M[\bar]{H} \geq \M{0}} \|\M{A}-\M{W}\M[\bar]{H}^\Tra\|$ is $n\times 2$ with $\M{C}=\M{A}^\Tra \M{W}$ and $\M{G}=\M{W}^\Tra\M{W}$}
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
\caption{Hierarchical Clustering \cite{KP13}}
\label{alg:hiernmf2}
\begin{algorithmic}[1]
	\Require{$\M{A}$ is $m\times n$, $k$ is target number of leaves}
	\Function{${\cal T} =$ Hier-R2-NMF}{$\M{A}$}
		\State ${\cal R} = \text{node}(\M{A})$ \hfill \Comment{create root node}
		\State \textsc{Split}$({\cal R})$
		\State inject$({\cal Q},{\cal R}.\text{left})$ \hfill \Comment{create priority queue}
		\State inject$({\cal Q},{\cal R}.\text{right})$ \hfill \Comment{of frontier nodes}
		\While{size$({\cal Q}) < k$}
			\State ${\cal N} = \text{eject}({\cal Q})$ \hfill \Comment{frontier node with max score}
			\State \textsc{Split}$({\cal N}.\text{left})$ \hfill \Comment{split left child}
			\State inject$({\cal Q},{\cal N}.\text{left})$ \hfill \Comment{and add to $\cal Q$}
			\State \textsc{Split}$({\cal N}.\text{right})$ \hfill \Comment{split right child}
			\State inject$({\cal Q},{\cal N}.\text{right})$ \hfill \Comment{and add to $\cal Q$}
		\EndWhile
	\EndFunction
	\Ensure{$\cal{T}$ is binary tree rooted at $\cal R$ with $k$ frontier nodes, each node has subset of cols of $\M{A}$ and feature vector $\V{w}$}
\end{algorithmic}
\end{algorithm}

\begin{figure}
\input{fig/tree}
\caption{Hierarchy node classification}
\label{fig:tree}
\end{figure}
